Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-02-10 src/tex/ms.tex
Update ms.tex
WIP updated Fig 1

\section{Introduction}
\label{sec:Introduction}
% Motivation
\Ac{GW} detections by the \ac{LVK} \citep{LIGO, Virgo, KAGRA} are now routinely used to test various aspects of Einstein's theory of \ac{GR} \citep{LIGOScientific:2016lio,LIGOScientific:2018dkp,LIGOScientific:2021sio}.
Among those, measurements of the basic properties of \acp{GW}, like their speed and polarization, can directly probe the fundamental symmetries of the underlying theory of gravity \citep{Will:2018bme}.
For instance, unequal propagation of \ac{GW} polarization states would reveal that spacetime is birefringent, a smoking gun for parity-odd theories like Chern-Simons gravity \citep{Jackiw:2003pm,Alexander:2009tp,Sopuerta:2009iy}.
Here we constrain the magnitude of possible amplitude birefringence using \ac{BBH} signals from the latest \ac{LVK} catalog, GWTC-3.

@@

% What's new?
In this study, we used a frequency-dependent birefringence model to constrain the strength of \ac{GW} amplitude birefringence by performing \ac{PE}.
This model is a better approximation of the birefringence effect than the frequency-independent model used in \citet{Okounkova_2022}.
Compared to previous studies, we also performed \ac{PE} on more events, including events in the third \ac{GW} transient catalog \citep{GWTC-3}, GWTC-3.
We consider 70 binary black hole merger events with a \ac{FAR} $\leq1\mathrm{yr^{-1}}$.
These events are also listed in Table I in \citet{GWTC-3_population}.\footnote{
GW190720 is included in the list in \citet{GWTC-3_population}, but we did not include it in this study.
Details in Sec.~\ref{sec:GW190720}.}
We then give the population constraint on the strength of \ac{GW} amplitude birefringence by using the \ac{PE} results.

@@

In particular, \emph{amplitude} birefringence would enhance one polarization mode over the other.
To first order in theories like Chern-Simons gravity, the Fourier-domain waveform observed a comoving distance $d_C$ away from the source takes the form of
\begin{equation}
    h_{L/R}^{\mathrm{br}}(f) =
    h_{L/R}^{\mathrm{GR}}(f) \times
    \exp\left(\pm\kappa\frac{d_C}{1\, \mathrm{Gpc}}\frac{f}{100\,\mathrm{Hz}}\right)\,,
    \label{eq:waveform_modification}
\end{equation}
where the emitted waveform $h_{L/R}^{\mathrm{GR}}$ is modified by an exponential birefringent factor to yield the observed waveform $h_{L/R}^{\mathrm{br}}$.
The overall magnitude of this effect for a given distance and frequency $f$ is set by a dimensionless opacity parameter, $\kappa$, which encodes the intrinsic strength of the birefringence.
The emitted waveform for a given source (i.e., the waveform observed very close to the source) will generally differ from the analogous waveform predicted by \ac{GR}; however, since we expect most viable modifications to \ac{GR} to be intrinsically small, it is standard to approximate the emitted waveform by the prediction from \ac{GR} (hence the notation ``$h^{\rm GR}$'' above).

@@

Equation \eqref{eq:waveform_modification} can be derived as the first order effect in an expansion for multiple extensions of \ac{GR}.
In general, $\kappa$ will be a function of the theory parameters and the cosmological history, e.g., the value of the pseudo-scalar field and its derivative in Chern-Simons gravity.
Since it originates from an expansion, Eq.~\eqref{eq:waveform_modification} is a good approximation only for small exponents, 
\begin{equation}
\left|\kappa\right| \left(d_C/1\,\mathrm{Gpc}\right) \left(f/100\, \mathrm{Hz}\right) < 1\, .
\end{equation}
Otherwise, more frequency-dependent terms would enter the exponent of Eq.~\eqref{eq:waveform_modification} in a theory-dependent way.
\mi{specify what the parameter for the expansion is, add citations, and check with a theorist}

@@

\citet{Wang_2021} formulated the birefringence effect as
\begin{equation}
    h_{+/\times}^{PV}(f) = h_{+/\times}^{GR}(f)\mp h_{\times/+}^{GR}(f)(i\delta h-\delta\Psi)\,,
\end{equation}
where $h_{+/\times}^{PV}(f)$ is the modified \ac{GW} waveform in the linear basis, $h_{+/\times}^{GR}(f)$ is the \ac{GW} waveform in the linear basis as \ac{GR} predicts, $\delta h$ is the amplitude modification, and $\delta\Psi$ is the phase modification.
We can rewrite the equation as
\begin{equation}
\begin{split}
    h_{L/R}^{PV}(f)&=h_{L/R}^{GR}(f)(1\mp\delta h\mp i\delta\Psi)\\
    &\approx h_{L/R}^{GR}(f)e^{(\mp\delta h\mp i\delta\Psi)}
\end{split}
\,,
\end{equation}
where $h_{L/R}^{PV}(f)$ is the modified \ac{GW} waveform in the circular basis, $h_{L/R}^{GR}(f)$ is the \ac{GW} waveform in the circular basis as \ac{GR} predicts.
This equation applies to any birefringence model that allows the assumption that the modification to \ac{GR} is small when observed very near the source.
To make a comparison with our work, we can choose $\delta\Psi=0$, which is a case that \citet{Wang_2021} considered.
In their work, they parametrized $\delta h=-A_\nu\pi f$, while
\begin{equation}
    A_\nu=M_{PV}^{-1}(\alpha_\nu(z=0)-\alpha_\nu(z)(1+z))\,,
\end{equation}
where $A_\nu$ is the parity-violating parameter for amplitude birefringence, $M_{PV}$ is the characteristic energy scale, and $\alpha_\nu$ is a arbitrary functions determined by the birefringence model.
They then chose $\alpha_\nu=1$.
Therefore, we can rewrite the equation as
\begin{equation}
    h_{L/R}^{PV}(f)= h_{L/R}^{GR}(f)\exp\left(\mp\frac{\pi zf}{M_{PV}}\right)\,.
\end{equation}
This equation takes the same form as \eqref{eq:waveform_modification}.
We can then find the relation between $M_{PV}$ and $\kappa$ by
\begin{equation}
    \left|\mp\frac{\pi zf}{M_{PV}}\right|=\left|\pm\kappa\frac{d_C}{1\,\mathrm{Gpc}}\frac{f}{100\,\mathrm{Hz}}\right|\,.
\end{equation}
Using $H_0d_C=cz$, where $H_0$ is the Hubble constant and $c$ is the speed of light, we can rewrite the equation as
\begin{equation}
    \kappa=\frac{\pi H_0}{c}\left(1\,\mathrm{Gpc}\right)\left(100\,\mathrm{Hz}\right)M_{PV}^{-1}\,.
\end{equation}

>>>>

\section{Introduction}
\label{sec:Introduction}
% Motivation
\Ac{GW} detections by the \ac{LVK} \citep{LIGO, Virgo, KAGRA} are now routinely used to test various aspects of Einstein's theory of \ac{GR} \citep{LIGOScientific:2016lio,LIGOScientific:2018dkp,LIGOScientific:2021sio}.
Among those, measurements of the basic properties of \acp{GW}, like their speed and polarization, can directly probe the fundamental symmetries of the underlying theory of gravity \citep{Will:2018bme}.
For instance, unequal propagation of \ac{GW} polarization eigenstates would reveal that spacetime is birefringent, a smoking gun for parity-odd theories like Chern-Simons gravity \citep{Jackiw:2003pm,Alexander:2009tp,Sopuerta:2009iy}.
Here we constrain the magnitude of possible amplitude birefringence using \ac{BBH} signals from the latest \ac{LVK} catalog, GWTC-3.

@@

% What's new?
In this study, we used a frequency-dependent birefringence model to constrain the strength of \ac{GW} amplitude birefringence by performing \ac{PE} on \ac{LVK} binaries.
This model is a better approximation of the birefringence effect than the frequency-independent model used in \citet{Okounkova_2022}.
Compared to previous studies, we also performed \ac{PE} on more events, including events in the third \ac{GW} transient catalog \citep{GWTC-3}, GWTC-3.
We consider 70 binary black hole merger events with a \ac{FAR} $\leq1\mathrm{yr^{-1}}$, as listed in Table I of \citet{GWTC-3_population}.
We use the results from individual events to place a collective population constraint on the strength of \ac{GW} amplitude birefringence from GWTC-3.

@@

In particular, \emph{amplitude} birefringence would enhance one polarization mode over the other.
To first order in theories like Chern-Simons gravity, the Fourier-domain waveform observed a comoving distance $d_C$ away from the source takes the form of
\begin{equation}
    h_{L/R}^{\mathrm{br}}(f) =
    h_{L/R}^{\mathrm{GR}}(f) \times
    \exp\left(\pm\kappa\frac{d_C}{1\, \mathrm{Gpc}}\frac{f}{100\,\mathrm{Hz}}\right)\,,
    \label{eq:waveform_modification}
\end{equation}
where the emitted waveform $h_{L/R}^{\mathrm{GR}}$ is modified by an exponential birefringent factor to yield the observed waveform $h_{L/R}^{\mathrm{br}}$.
The overall magnitude of this effect for a given distance and frequency $f$ is set by a dimensionless opacity parameter, $\kappa$, which encodes the intrinsic strength of the birefringence.
The emitted waveform for a given source (i.e., the waveform observed in the near zone, very close to the source) will generally differ from the analogous waveform predicted by \ac{GR}; however, since we expect most viable modifications to \ac{GR} to be intrinsically small, it is standard to approximate the emitted waveform by the prediction from \ac{GR} (hence the notation ``$h^{\rm GR}$'' above).

@@

Equation \eqref{eq:waveform_modification} can be derived as the first order effect in an expansion away from \ac{GR} under multiple frameworks.
In general, $\kappa$ will be a function of the theory parameters and the cosmological history, e.g., the value of the pseudo-scalar field and its derivative in Chern-Simons gravity.
Since it originates from a truncated series expansion, Eq.~\eqref{eq:waveform_modification} is a good approximation only for small exponents, 
\begin{equation}
\left|\kappa\right| \left(d_C/1\,\mathrm{Gpc}\right) \left(f/100\, \mathrm{Hz}\right) < 1\, .
\end{equation}
Otherwise, more frequency-dependent terms would enter the exponent of Eq.~\eqref{eq:waveform_modification} in a theory-dependent way.
\mi{specify what the parameter for the expansion is, add citations, and check with a theorist}

@@

\citet{Wang_2021} formulated the birefringence effect as
\begin{equation}
    h_{+/\times}^{PV}(f) = h_{+/\times}^{GR}(f)\mp h_{\times/+}^{GR}(f)(i\delta h-\delta\Psi)\,,
\end{equation}
where $h_{+/\times}^{PV}(f)$ is the modified \ac{GW} waveform in the linear basis, $h_{+/\times}^{GR}(f)$ is the \ac{GW} waveform in the linear basis as \ac{GR} predicts, $\delta h$ is the amplitude modification, and $\delta\Psi$ is the phase modification.
We can rewrite the equation as
\begin{equation}
\begin{split}
    h_{L/R}^{PV}(f)&=h_{L/R}^{GR}(f)(1\mp\delta h\mp i\delta\Psi)\\
    &\approx h_{L/R}^{GR}(f)\exp\left(\mp\delta h\mp i\delta\Psi\right)
\end{split}
\,,
\end{equation}
where $h_{L/R}^{PV}(f)$ is the modified \ac{GW} waveform in the circular basis, $h_{L/R}^{GR}(f)$ is the \ac{GW} waveform in the circular basis as \ac{GR} predicts.
This equation applies to any birefringence model that allows the assumption that the modification to \ac{GR} is small when observed very near the source.
To make a comparison with our work, we can choose $\delta\Psi=0$, which is a case that \citet{Wang_2021} considered.
In their work, they parametrized $\delta h=-A_\nu\pi f$, while
\begin{equation}
    A_\nu=M_{PV}^{-1}(\alpha_\nu(z=0)-\alpha_\nu(z)(1+z))\,,
\end{equation}
where $A_\nu$ is the parity-violating parameter for amplitude birefringence, $M_{PV}$ is the characteristic energy scale, and $\alpha_\nu$ is a arbitrary functions determined by the birefringence model.
They then chose $\alpha_\nu=1$.
Therefore, we can rewrite the equation as
\begin{equation}
    h_{L/R}^{PV}(f)= h_{L/R}^{GR}(f)\exp\left(\mp\frac{\pi zf}{M_{PV}}\right)\,.
\end{equation}
This equation takes the same form as \eqref{eq:waveform_modification}.
We can then find the relation between $M_{PV}$ and $\kappa$ by
\begin{equation}
    \left|\mp\frac{\pi zf}{M_{PV}}\right|=\left|\pm\kappa\frac{d_C}{1\,\mathrm{Gpc}}\frac{f}{100\,\mathrm{Hz}}\right|\,.
\end{equation}
Using $H_0d_C=cz$, where $H_0$ is the Hubble constant and $c$ is the speed of light, we can rewrite the equation as
\begin{equation}
    \kappa=\frac{\pi H_0}{c}\left(1\,\mathrm{Gpc}\right)\left(100\,\mathrm{Hz}\right)M_{PV}^{-1}\,.
\end{equation}
